---
title: "Exploring Alpha Priors in Dirichlet Process"
author: "Jose Calatayud Mateu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this document, we explore the effects of different prior distributions on the alpha parameter in a Dirichlet Process Gaussian Mixture Model. We begin by generating synthetic data from a lognormal mixture model, fitting a Dirichlet process model, and then analyzing how different alpha priors influence the number of clusters formed.

## Simulating and Fitting a Lognormal Mixture Model

We first generate data from a mixture of two lognormal distributions to simulate a complex dataset that may require flexible clustering. We then fit a Dirichlet Process Gaussian Mixture Model to this data.

```{r}
set.seed(42)

# Generate synthetic data from a lognormal mixture model
data1 <- rlnorm(100, meanlog = 1, sdlog = 0.3)  
data2 <- rlnorm(100, meanlog = 2, sdlog = 0.3)  
simulated_data <- c(data1, data2)  

# Fit Dirichlet Process Gaussian Mixture Model
dp_lognormal <- DirichletProcessGaussian(simulated_data)
dp_lognormal <- Fit(dp_lognormal, 1000)  

# Graphics
plot(dp_lognormal)
```

```{r}
# Define a function from the posterior
posterior_func <- PosteriorFunction(dp_lognormal)

# Generate new samples by evaluating the function on random inputs
new_samples <- posterior_func(simulated_data)

# Calculate 5% and 95% quantiles
quantiles <- quantile(new_samples, probs = c(0.05, 0.95))
print(quantiles)
```

We generate two lognormal distributions with different means and standard deviations. Then, the data is fitted using a Dirichlet Process Gaussian Mixture Model, allowing flexible clustering, so we run 1000 iterations to refine the model and sample new data from the posterior. We summarize the 5% and 95% quantiles of the posterior samples to understand the spread of the modeled distribution.

## Exploring the Effect of Alpha Prior

The alpha parameter in a Dirichlet Process controls the tendency of the model to create new clusters. A low alpha encourages fewer clusters, while a high alpha promotes more clusters. Here, we compare models with different priors on alpha.

```{r alpha_effect}
# Processes
dp_alpha1 <- DirichletProcessGaussian(simulated_data) #default
dp_alpha2 <- DirichletProcessGaussian(simulated_data, 
                                      alphaPriors = c(1,2)) # low alpha
dp_alpha3 <- DirichletProcessGaussian(simulated_data, 
                                      alphaPriors = c(6,2)) # high alpha

# Fitting model
dp_alpha1 <- Fit(dp_alpha1, 1000)
dp_alpha2 <- Fit(dp_alpha2, 1000)
dp_alpha3 <- Fit(dp_alpha3, 1000)

# Extract alpha parameter chains
alpha_chain1 <- dp_alpha1$alphaChain
alpha_chain2 <- dp_alpha2$alphaChain
alpha_chain3 <- dp_alpha3$alphaChain

# Plot alpha chains to observe convergence
iterations <- seq_along(alpha_chain1)
data_plot <- data.frame(
  Iteration = rep(iterations, 3),
  Alpha = c(alpha_chain1, alpha_chain2, alpha_chain3),
  Prior = rep(c("Alpha = (2,4)", "Alpha = (1,2)", "Alpha = (6,2)"), each = length(iterations))
)

# Plot using ggplot2
ggplot(data_plot, aes(x = Iteration, y = Alpha, color = Prior)) +
  geom_line(linewidth = 0.3) + 
  labs(title = "Alpha Parameter Chains", x = "Iterations", y = "Alpha Value") +
  theme_minimal()
```

Here, we define three different alpha priors to test their effect on clustering. Each model is fitted with 1000 iterations to allow proper convergence. We extract the alpha parameter chains to visualize how the parameter evolves during fitting. The plot compares the different alpha chains, helping assess how quickly each model converges and how stable the estimates are.

## Conclusion

This analysis demonstrates that the choice of prior on the alpha parameter significantly affects the number of clusters formed in a Dirichlet Process Gaussian Mixture Model. Lower alpha values lead to fewer clusters, while higher values promote more clusters. By examining the alpha parameter chains, we also gain insights into the convergence behavior of the model under different priors.
